---
title: "Quantifying Diversity"
author: "Yuval Mendelson"
date: "11-03-2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this notebook I would try to quantify the diversity of the fish surveys in the project.

```{r}
library(vegan)
library(tidyverse)
library(plotrix)
library(betapart)
```

Load the data:

```{r}
orig_df <- readRDS("./EilatFishCommunityEcology/Data/wide_fish_eilat.rds")
# make data table into data frame:
orig_df <- orig_df %>% ungroup()
orig_df <- as.data.frame(orig_df)


first_species_col = 12

zero_trans_removed <- apply(sp_matrix, 1, function(row) sum(as.numeric(row[first_species_col:length(row)])) != 0)
no_zeros_trans_df <- orig_df[zero_trans_removed,]
head(no_zeros_trans_df)
```

lets make the species matrix:

```{r}
sp_matrix <- no_zeros_trans_df[,first_species_col:length(orig_df)]

head(sp_matrix)
```

### Diversity Profiling - Hills numbers

first, I would like to calculate the Hills numbers for each site in the data.

```{r}
renyi_profile <- renyi(
  sp_matrix,
  scales = c(0, 0.05, 0.1, 0.2, 0.4, 0.6, 0.8, 1, 2, 4, 8, 16, 32, 64,            Inf),
  hill = T
)

#renyi_profile[is.na(renyi_profile)] <- 0
head(renyi_profile)
```

now we want to save some of the metadata columns from the original data:

```{r}
metadata_df <- no_zeros_trans_df[1:first_species_col-1]
renyi_df <- bind_cols(metadata_df, renyi_profile)
```

now I want to create the plot:

```{r}
renyi_df <- gather(renyi_df, "Q", "Value", (first_species_col-1):length(renyi_df))
renyi_df$Q <- as.numeric(renyi_df$Q)

ggplot( data = renyi_df ) +
  aes(x = Q, y = Value, group = survey_id , color = Site)+
  geom_line() +
  geom_point()+
  scale_x_continuous(limits = c(0,64))
```

#### Site specific:

```{r}
for (site in unique(renyi_df$Site))
{
  p <- renyi_df %>% 
    filter(Site == site) %>% 
    ggplot() +
    aes(x = Q, y = Value, group = survey_id , color = Year)+
    geom_line() +
    geom_point()+
    scale_x_continuous(limits = c(0,64)) +
    ggtitle(paste("Site - ", site))
  print(p)
}
```

## TODO: make more plots based on season-year and site specific

## Beta Diversity
