---
title: "QuantifyingDiversity"
author: "Yuval Mendelson"
date: "2024-03-05"
output: html_document
---

# Overview

In this markdown I would like to do the quantified diversity of the project.
Diversity if the number of species in an area ans their abundance.
first we need to load the needed libraries:
```{r}
library(vegan)
library(tidyverse)
library(plotrix)
library(betapart)
```

now we need to load the data:
```{r}
m_df <- readRDS("wide_fish_eilat.rds")
first_species_col = 12
sp_matrix <- m_df[,first_species_col:ncol(m_df)]
```

## Diversity Profile - Hill Numbers
```{r}
renyi_profile <- renyi(sp_matrix, 
                       scales = c(0, 0.05, 0.1, 0.2, 0.4, 0.6, 0.8, 1.0, 2.0, 
                                  4.0, 8.0, 16.0, 32.0, 64.0, Inf),
                       hill=T)
```
now we would like to add to the reyni matrix some meta data:
```{r}
meta_data <- m_df[,1:10]
renyi_df <- bind_cols(meta_data, renyi_profile)
```
now we would like to arrange the data frame a bit:
```{r}
renyi_df <- gather(renyi_df, "Q", "value", 11:25)
#renyi_df <- renyi_df %>% arrange(Site_name, Year)
renyi_df$Q <- as.numeric(renyi_df$Q)

renyi_df %>% 
  filter(Site_name == "Caves") %>% 
  ggplot() +
  aes(x = Q, y = value, group = trans_ID, color = Site_name) + 
  geom_line() +
  geom_point() 
```
now for each site seperatly:
```{r}
for (val in unique(renyi_df$Site_name))
{
  p <- renyi_df %>% 
    filter(Site_name == val) %>% 
    ggplot() +
    aes(x = Q, y = value, group = trans_ID) +
    geom_line() +
    geom_point() +
    ggtitle(paste("Hills number of ", val))
  print(p)
}
```

